---
title: "Population, Births, Deaths"
author: "Edyta Pszczółkowska 435022"
date: "8 01 2022"
output: html_document
---


###  Introduction


We will cover  4 main topics regarding living in Barcelona: 

* Population, births, deaths (Edyta)
* Unemployment (Edyta)
* Transport, accidents (Min)
* Immigration (Min)



Our main question which we want to answer by following visualisations is:

***Is Barcelona a good place to live?***



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE )

```

# Loading necessary packages and datasets

The whole dataset consists of 17 separate csv files.


```{r}

library(ggplot2)
library(dplyr)
library(scales)
library(tidyr)
library(RColorBrewer)
library(ggcharts)
library(ggthemes)
library(ggplot2)
library(ggrepel)
library(tidyverse)

setwd("D:\\Edyta pliki\\DS\\III semestr\\AdvVis in R\\project")

#accidents <- read.csv("accidents_2017.csv")
#air_quality_Nov <- read.csv("air_quality_Nov2017.csv")
#air_stations_Nov <- read.csv("air_stations_Nov2017.csv")
births <- read.csv("births.csv")
#bus_stops <- read.csv("bus_stops.csv")
deaths <- read.csv("deaths.csv")
#immigrants_by_nationality <- read.csv("immigrants_by_nationality.csv")
immigrants_emigrants_by_age <- read.csv("immigrants_emigrants_by_age.csv")
#immigrants_emigrants_by_destination <- read.csv("immigrants_emigrants_by_destination.csv")
#immigrants_emigrants_by_destination2 <- read.csv("immigrants_emigrants_by_destination2.csv")
immigrants_emigrants_by_sex <- read.csv("immigrants_emigrants_by_sex.csv")
#life_expectancy <- read.csv("life_expectancy.csv")
#most_frequent_baby_names <- read.csv("most_frequent_baby_names.csv")
population <- read.csv("population.csv")
#transports <- read.csv("transports.csv")
#unemployment <- read.csv("unemployment.csv")
```


### Population

Let us plot the distribution of age and gender in the whole population with the most apropriate tool to achieve it: pyramid chart:

```{r}
population2017 <- population %>%
  filter(Year=="2017") %>%
  group_by(Age, Gender) %>%
  summarise(Count = sum(Number))


population2017$Age[population2017$Age == "5-9"] <- "05-09"
population2017$Age[population2017$Age == ">=95"] <- "95+"
pyramid_chart(data = population2017, x = Age, y = Count, group = Gender, title = "Population pyrimid as of 2017", xlab = "Number" ) 

```


We can see, that the demographic situation in Barcelona is rather stable one: there is enough children and youngsters to replace current labour force in the future.


## Births grouped by years


```{r}
# Births by year
births %>%
  group_by(Year, Gender) %>%
  summarise(Count=sum(Number)) %>%
  mutate(percent=paste0(round((Count/sum(Count))*100, 2), "%")) %>%
  ggplot(aes(x=Year, y=Count, fill=Gender)) +
  geom_bar(stat="identity", position="stack") + 
  geom_text(aes(label=percent, group=Gender), position=position_stack(vjust=0.5)) +
  labs(x="Year", y="Births", title="Births by year (2013-2017)") +
  scale_y_continuous(labels=comma, breaks = seq(0,15000, 2500)) +
  theme_bw()

```


The number of children born in years 2013-2017 remain on the same level. It is quite puzzling, why it were always boys that were born more than girls.

### Population by district

```{r}
population_by_district <- population %>%
  filter(Year == "2017") %>%
  group_by(District.Name) %>%
  summarise(Pop_count = (sum(Number))) %>%
  arrange(desc(Pop_count))  

# population_by_district
# 
# ggplot(population_by_district, aes(x="", y=Pop_count, fill=District.Name)) +
#   geom_bar(stat="identity", width=1, color="white") +
#   coord_polar("y", start=0)+
#   
#   theme_void()

data_for_pie <- population_by_district %>% 
  arrange(desc(District.Name)) %>%
  mutate(prop = Pop_count / sum(population_by_district$Pop_count) *100) %>%
  mutate(ypos = cumsum(prop)- 0.5*prop ) %>%
  arrange((Pop_count))

data_for_pie$value <-  (round(data_for_pie$prop,2)) 



df2 <- data_for_pie %>% 
  mutate(csum = rev(cumsum(rev(value))), 
         pos = value/2 + lead(csum, 1),
         pos = if_else(is.na(pos), value/2, pos))

ggplot(df2, aes(x = "" , y = value, fill = fct_inorder(District.Name))) +
  geom_col(width = 1, color = 1) +
  coord_polar(theta = "y") +
  scale_fill_brewer(palette = "Oranges") +
  geom_label_repel(data = df2,
                   aes(y = pos, label = paste0(value, "%")),
                   size = 4.5, nudge_x = 1, show.legend = FALSE) +
  guides(fill = guide_legend(title = "District Names")) +
  theme_void() +   
  labs(title =  "Percent of population with respect to districts",
       subtitle = '2017')
```





## Immigrants in Barcelona


Barcelona is a place which seems to be a paradise to live: warm pleasant mediterranean climat, culture, food and atmosphere.
It is a very popular travel destination. But how many people decide to stay there longer:

The following graph gives us an answer.
Let's check how big is the ratio of immigrants to the whole population:


```{r}

immigrants_2013_2017  <- immigrants_emigrants_by_sex %>%
                        group_by(Year) %>%
                        summarise(Immigrants = sum(Immigrants))




  

immigrants_2013_2017

```

```{r}
population_2013_2017 <- population  %>%
  group_by(Year) %>%
  summarise(Count_population = sum(Number))
population_2013_2017
```


```{r}
pop_immigrants <- merge(immigrants_2013_2017,population_2013_2017 , by = "Year" )

pop_immigrants$ratio <- round(pop_immigrants$Immigrants / pop_immigrants$Count_population * 100 ,2 )

pop_immigrants$plot_label <- paste0(pop_immigrants$ratio , "%")

pop_immigrants
```


```{r}
ggplot(pop_immigrants) +
   geom_bar(aes(x =  Year , y = Immigrants) , stat = "identity")  +
  geom_line(aes(x =  Year, y = ratio*10000, color = "ratio of immigrants in the whole population [%]" ), size= 2 ) + 
  theme(legend.position = 'bottom')  +
  scale_color_manual(values = c('red'))  +
   labs(color = NULL)  +
     scale_y_continuous( 
    labels = comma,
    sec.axis = sec_axis(~./5000),  )   + 
  labs(title =  "Immigrants in absolute terms and as part of population , 2013-2017")

```





