---
title: "Edyta Pszczolkowska (ID 435022) - association rules using apriori algorithm"
output:
  html_document:
    df_print: paged
---

In this paper I will perform apriori algorithm on a dataset of transactions from a bakery. 
the aim of such a project is to discover rules indicating what drives consumers to buy 
certain products or the other way round: what are they likely to
buy additionally once they put particular set of items to their basket.

```{r}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

```{r}

#loading packages and preparing dataset

library(arulesViz)
library(arules)

df <- read.csv("bread basket.csv")

df <- df[ , c(1,2)]

dir.create(path = "tmp", showWarnings = FALSE)

# Write our data.frame to a csv
write.csv(df, "./tmp/my_file.csv")

# Read that csv back in
bakery <- read.transactions(
  file = "./tmp/my_file.csv",
  format = "single",
  sep = ",",
  cols=c("Transaction","Item"),
  header = TRUE,
  rm.duplicates = T
)
summary(bakery)

```

```{r}
#SUPPORT - one of three most basic measures in market basket analysis
itemFrequency(bakery, type="relative")
itemFrequency(bakery, type="absolute")



#Let's plot the mosr frequent ones:
itemFrequencyPlot(bakery, support = 0.1)

itemFrequencyPlot(bakery, topN=10, type="absolute", main="Items Frequency") 

#Let's plot 100 random transactions, on the X-axis there are numbers representing particular items, those
#products which are often purchased form a vertical line on the plot.
image(sample(bakery, 100))
##################################################
#LETS PLAY A GAME!!!

#The choice of parameters is arbitrary, but there is a trade off between quality of measures and  minimum length of a dataset, 
#if one of them is high, then the second one need to be adequately lower.
bakeryrules <- apriori(bakery, parameter = list(support = 0.007, confidence = 0.01, minlen = 3))

#Following commands define the most important rules sorted by three major measures: support, lift and confidence. 
inspect(sort(bakeryrules, by = "lift")[1:5])
inspect(sort(bakeryrules, by = "support")[1:5])
inspect(sort(bakeryrules, by = "confidence")[1:5])

summary(bakeryrules)



# what drives people to buy coke?
rules.coke<-apriori(data=bakery, parameter=list(supp=0.01,conf = 0.005), 
                       appearance=list(default="lhs", rhs="Coke"), control=list(verbose=F)) 
rules.coke.byconf<-sort(rules.coke, by="confidence", decreasing=TRUE)
inspect(head(rules.coke.byconf))

#as we can see customers the most often buy coke without anything - maybe simply a promotion 2 coke +1 for free would be a good idea?

# on the opposite - what do customers prefer when they eat sandwich: tea, coffee, a juice...?
rules.sandwich<-apriori(data=bakery, parameter=list(supp=0.005,conf = 0.005), 
                          appearance=list(default="rhs", lhs="Sandwich"), control=list(verbose=F)) 
rules.sandwich.byconf<-sort(rules.sandwich, by="confidence", decreasing=TRUE)
inspect(head(rules.sandwich.byconf))

#they are more likely to buy a coffee - let's offer a sandwich + coffee for 1€

#Let's check significance of the rules with Fisher’s exact :
is.significant(rules.coke, bakery)
is.significant(rules.sandwich, bakery)

#Redundancy - if there is a more general one with the same or higher confidence value.
is.redundant(rules.coke)
is.redundant(rules.sandwich)




library(arulesCBA)

#Let's plot now all three major indicators at once:

plot(bakeryrules, measure=c("support","lift"), shading="confidence")
plot(bakeryrules, method="grouped")
plot(bakeryrules, method="graph", control=list(type="items"))

```


CONCLUSION

In my personal opinion, the apriori algorithm is not complicated, but the insight we obtain from its output can give us extremely useful
knowledge. Moreover it has much broader application than only market basket analysis, e.g. based on particular traits we can predict a result of
an experiment, survey.
